Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-03-26 GWLDCI.tex
added intro and bibliographical stubs

***Some general intro to WLDs perhaps similar to the end of p2 and top of p3 of \cite{wilsonloop}.  For this paper we don't need the integral, so for the moment they are not defined below, but for motivation may want to be glossed a bit here ***

Furthermore, in \cite{wilsonloop}, Agarwala and Amat show that Wilson loop diagrams can be interpreted as positroids, a certain well behaved class of realizable matroids.  This opens up the study of Wilson loop diagrams to techniques from geometry and combinatorics.  We pursue this direction here and in the companion paper ***cite the other paper***.

In the present paper we focus on the combinatorics and geometry of an equivalence relation on Wilson loop diagrams.  The main results are that Wilson loop diagrams are equivalent if and only if they define the same matroid, that the equivalence can be rephrased in terms of retriangulations in polygon dissections, and from there that the sizes of the equivalence classes can be enumerated, and the set of equivalence classes is related to directions in associahedra.

@@

We proceed in Section~\ref{section overall background} with the required definitions for Wilson loop diagrams (subsection~\ref{section WLD background}) and matroids (subsection~\ref{sec matroid background}.  Wilson loop diagrams will be said to be admissible when they correspond to the physical situation of ***something***, or mathematically according to Definition~\ref{admisdfn}.  With this, the key definition is Definition~\ref{equivdfn} giving the equivalence relation on admissible Wilson loop diagrams.  The connection to matroids is stated precisely in Theorem \ref{thm WLD defines matroid}.

@@

We depict this data as a circle with marked points, called vertices. The vertices are labeled by $V$ (preserving the cyclic ordering). The arc between consecutive vertices are called edges. There are $k$ wavy lines in the interior of the diagram, depicting the proagators, with endpoints on the edges. A propagator, $p =(i,j)$ has one endpoint on the edge between the vertex labeled $i$ and $i+1$ and another endpoint on the edge defined by $j$ and $j+1$. The condition on $i_r$ and $j_r$ means that the propagator does not go between adjacent edges. Let $\cP = \{p_r\}_{r=1}^k$ be the set of propagators. Then we write \bas W = (\cP, V) \;.\eas

@@

\begin{dfn}
  Let $W = (\cP, [n])$ be an admissible Wilson loop diagram.  The \emph{polygon dissection} associated to $W$, denoted $\tau(W)$, is defined as follows.
  \begin{itemize}
  \item The vertices of $\tau(W)$ correspond to the edges of $W$.
  \item Labeling the vertices of $\tau(W)$ with the edge number of $W$gives a cyclic order to the vertices. Connecting consecutive vertices gives a graph theoretic cycle called the polygon of $\tau(W)$.
  \item Each propagator of $W$ defines a chord edge of $\tau(W)$; specifically,  a propagator $(i,j) \in \cP$ defines a chord connecting the vertices $i$ and $j$ in $\tau(W)$.
  \end{itemize}
\end{dfn}

@@

The connection between admissible Wilson loop diagrams and polytope dissections suggests a connection to associahedra. First we need some background.

@@

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{}; the following will be the one which we will use.

@@

With the choice of $n$-gon and hence of $A_n$ fixed, view the $\tau$ map as taking values in the set of dissections of this particular convex $n$-gon, and so each weakly admissible Wilson loop diagram is associated via $\tau$ with a face (in some dimension) of $A_n$.

@@

\begin{proof}
  Let $W$ and $X$ be equivalent.  From results in the paper we're working on we know that $\tau(W)$ and $\tau(X)$ differ from each other only by retriangulations of their triangulated pieces.

@@

  Consider now an edge whose quadrilateral has corners $x_i$, $x_j$, $x_k$, and $x_l$ in that order.  Let $t_1$ and $t_2$ be the triangulations corresponding to the two ends of the edge where $t_1$ has a diagonal from $x_j$ to $x_l$ and $t_2$ has a diagonal from $x_i$ to $x_k$.  From the definition of $A_n$ we see that $s_{t_1}$ and $s_{t_2}$ only differ at coordinates $i$, $j$, $k$, and $l$, and all of the triangles incident to any of these vertices other than the one coming from triangulating $x_i, x_j, x_k, x_l$ are the same in both $t_1$ and $t_2$.  Let $a$ be the area of the quadrilateral and let $a(i,j,k)$ be the area of the triangle with corners $x_i$, $x_j$ and $x_k$, and similarly for other indices then $s_{t_1}$ on coordinates $i$, $j$, $k$, $l$ is $(a(i,j,l) + b_i, a+b_j, a(k,l,j)+b_k, a+b_l)$ and $s_{t_2}$ on these coordinates is $(a + b_i, a(j,k,i)+b_j, a+b_k, a(l,i,k)+b_l$, where the $b_i$ are the sums of the areas of the relevant triangles which are common to both triangulations.  Thus a vector parallel to this edge is the vector which is $(a(i,j,l) - a, a-a(j,k,i), a(k,l,j)-a, a-a(l,i,k)) = (-a(j,k,l), a(i,k,l), -a(i,j,l), a(i,j,k))$ on $i$, $j$, $k$, $l$ and $0$ in all other coordinates.

@@

The following lemma is useful for the second proposition.

\begin{lem}\label{lem good quads}
  Let $D_1$ and $D_2$ be two dissections of the convex polygon defined by the vertices $x_1, \ldots, x_n$.  Suppose the vertices incident to at least one non-triangle face are the same in both dissections, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ such that in one of $D_1$ or $D_2$ all four vertices are on the same non-triangle face while in the other there is a diagonal of the dissection such that there is one side of the diagonal with exactly one of $\{x_i, x_j, x_k, x_l\}$ (so the others are end points of this diagonal or strictly on the other side).
\end{lem}

@@

  Suppose this never occurs.  Then all the non-triangle faces of $D_2$ which are crossed by a diagonal of $D_1$ must be quadrilaterals and likewise for $D_1$.  This means that any non-triangle and non-quadrilateral faces must be common between $D_1$ and $D_2$.  These common faces are irrelevant for the statement, and so by triangulating any common non-triangle faces, we may assume that $D_1$ and $D_2$ have only quadrilateral and triangle faces.

>>>>

This paper is the first in a two part series investigating the combinatorics and geometry underlying SYM $N=4$ theory. This series lays out several results about the relationship between the Wilson loop diagrams and the positroid cells that give a CW complex structure to the positive Grassmanian. This paper expands on an observation from \cite{wilsonloops}, that shows that certain families of Wilson loop diagrams correspond to the same positroid cell. This paper gives a way of enumerating the number of different Wilson loop diagrams that correspond to a given positroid cell. Furthermore, it shows that that the number of cells in the image of the map to positive Grassmannians is given by counting the directions of associahedrons.

The second paper in this series identifies which positroid cell is associated to which Wilson loop diagram \cite{GWLDCII}. It proves that a Wilson loop diagram with $k$ propagators corresponds to a $3k$ dimensional positroid cell. Furthermore, it gives an algorithm for passing from a Wilson loop diagram to the Grassman necklace that identifies the associated positroid cell. In this manner the authors give a means of identifying which positroid cell is associate to which Wilson loop diagram that circumvents the painstaking process of examining bases sets. Finally, the authors characterize the the volume forms that Wilson loop diagrams associate to each positroid cells.

In recent years, there has been a an active program researching the geometric and combinatorics underlying SYM $N=4$ theory \cite{wilsonloops, amplituhedron, correlahedron, karpwilliams, galashinlam}. This body of work started with the observation that the on shell amplitudes of this theory correspond to the volume of a subspace of a positive Grasmannian, called an Amplituhedron \cite{amplituhedron}. Since then, there has been significant work to study the structure of the amplituhedron, geometrically and combinatorially, for instance \cite{amplituhedron2, karpwilliams, galashinlam}.

Meanwhile, there has been a different body of work studying SYM $N=4$ theory from the point of view of Feynman integrals \cite{BullimoreSkinner, AldayMaldecena, others}. The first author and Amat uncovered a connection between the Feynman integrals developed in this literature and the positive Grassmannians \cite{wilsonloops}. Others have tried to define a geometric space associated to these integrals in a manner similar to the Amplituhedron \cite{correlahedron, non-orientable}.

In this paper, we expand on map between Wilson loop diagrams and positroids, first defined in \cite{wilsonloops}. Each Wilson loop diagram, $W$, has associated to it a real-variable valued matrix $C(W)$, which defines positroid, i.e. a matroid that can be realized as an element of $\Gr(k,n)$. The subspace of $\Gr(k,n)$ parametrized by each $C(W)$ is the positroid cell defined by $W$, called $\Sigma(W)$. Lemma ?? of \cite{wilsonloops} identifies that certain Wilson loop diagrams define the same positroid cells. In particular, the authors define a subset of Wilson loop diagrams, called exact diagrams, and show that if two Wilson loop diagrams only differ by an exact subdiagram, they define the same positroid cell.

In the current work we study exact subdiagrams in greater detail, identifying several matroidal properties. Furthermore, we enumerate how many different Wilson loop diagrams map to any positroid cell, $\Sigma(W)$. Furthermore, we show that counting the number of distinct $\Sigma(W)$, positroid cells that correspond to Wilson loop diagrams, corresponds to counting the number of directions of an associahedron.

@@

We proceed in Section~\ref{section overall background} with the required definitions for Wilson loop diagrams (subsection~\ref{section WLD background}) and matroids (subsection~\ref{sec matroid background}.  Wilson loop diagrams  are said to be admissible when they correspond to $N^kMHV$ diagrams, or mathematically according to Definition~\ref{admisdfn}.  With this, the key definition is Definition~\ref{equivdfn} giving the equivalence relation on admissible Wilson loop diagrams.  The connection to matroids is stated precisely in Theorem \ref{thm WLD defines matroid}.

@@

We depict this data as a circle with marked points, called vertices. The vertices are labeled by $V$ (preserving the cyclic ordering). The arc between consecutive vertices are called edges. There are $k$ wavy lines in the interior of the diagram, depicting the proagators, with endpoints on the edges. A propagator, $p =(i,j)$ has one endpoint on the edge between the vertex labeled $i$ and $i+1$ and another endpoint on the edge defined by $j$ and $j+1$. The condition on $i_r$ and $j_r$ means that the propagator does not go between adjacent edges. Let $\cP = \{p_r\}_{r=1}^k$ be the set of propagators. Then we write \bas W = (\cP, V) \;.\eas \sanote{picture}

@@

\begin{dfn}\label{WLDtriangulationdfn}
  Let $W = (\cP, [n])$ be an admissible Wilson loop diagram.  The \emph{polygon dissection} associated to $W$, denoted $\tau(W)$, is defined as follows.
  \begin{itemize}
  \item The vertices of $\tau(W)$ correspond to the edges of $W$.
  \item Labeling the vertices of $\tau(W)$ with the edge number of $W$gives a cyclic order to the vertices. Connecting consecutive vertices gives a graph theoretic cycle called the polygon of $\tau(W)$.
  \item Each propagator of $W$ defines a chord edge of $\tau(W)$; specifically,  a propagator $(i,j) \in \cP$ defines a chord connecting the vertices $i$ and $j$ in $\tau(W)$.
  \end{itemize}
\end{dfn}

@@

The connection between admissible Wilson loop diagrams and polytope dissections suggests a connection to associahedra. First we need some background. \sanote{expand}

@@

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{}; the following will be the one which we will use. \sanote{picture of an associahedron somewhere? or do people know what it is?}

@@

The map $\tau$ defined in Definition \ref{WLDtriangulationdfn} assigns a dissection of an $n$-gon to a Wilson loop diagram $W = (\cP, n)$. Recall that if $|\cP| = k$, then $\tau(W)$ is a dissection with $k$ chords. Therefore, the map $\tau$ associates to each weakly admissible Wilson loop diagram a $n-3 - k$ dimensional face of the associahedron $A_n$. \sanote{picture of $\tau(W)$ for $k =1$, $n = 5$?}

@@

\begin{proof}
  Let $W$ and $X$ be two equivalent Wilson loop diagrams.  From \hlfix{Lemma {\ref{??}}}{Do we actually have this explicitly? Feels like we should}, we know that $\tau(W)$ and $\tau(X)$ differ from each other only by retriangulations of their triangulated pieces.

@@

  Consider now an edge whose quadrilateral has corners $x_i$, $x_j$, $x_k$, and $x_l$ in that order.  Let $t_1$ and $t_2$ be the triangulations corresponding to the two ends of the edge where $t_1$ has a diagonal from $x_j$ to $x_l$ and $t_2$ has a diagonal from $x_i$ to $x_k$.  From the definition of $A_n$ we see that $s_{t_1}$ and $s_{t_2}$ only differ at coordinates $i$, $j$, $k$, and $l$, and all of the triangles incident to any of these vertices other than the one coming from triangulating $x_i, x_j, x_k, x_l$ are the same in both $t_1$ and $t_2$.  Let $a$ be the area of the quadrilateral and let $a(i,j,k)$ be the area of the triangle with corners $x_i$, $x_j$ and $x_k$, and similarly for other indices then $s_{t_1}$ on coordinates $i$, $j$, $k$, $l$ is $(a(i,j,l) + b_i, a+b_j, a(k,l,j)+b_k, a+b_l)$ and $s_{t_2}$ on these coordinates is $(a + b_i, a(j,k,i)+b_j, a+b_k, a(l,i,k)+b_l$, where the $b_i$ are the sums of the areas of the relevant triangles which are common to both triangulations.  Thus a vector parallel to this edge is the vector which is $(a(i,j,l) - a, a-a(j,k,i), a(k,l,j)-a, a-a(l,i,k)) = (-a(j,k,l), a(i,k,l), -a(i,j,l), a(i,j,k))$ on $i$, $j$, $k$, $l$ and $0$ in all other coordinates. \sanote{definitely need a picture}

@@

The following lemma is useful for proving the converse of the above.

\begin{lem}\label{lem good quads}
  Let $D_1$ and $D_2$ be two dissections of the convex polygon defined by the vertices $x_1, \ldots, x_n$.  \hlfix{Suppose the vertices incident to at least one non-triangle face are the same in both dissections}{relevant?}, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ such that in one of $D_1$ or $D_2$ all four vertices are on the same non-triangle face while in the other there is a diagonal of the dissection such that there is one side of the diagonal with exactly one of $\{x_i, x_j, x_k, x_l\}$ (so the others are end points of this diagonal or strictly on the other side).
\end{lem} \sanote{picture?}

@@

Suppose this never occurs.  Then all the non-triangle faces of $D_2$ which are crossed by a diagonal of $D_1$ must be quadrilaterals and likewise for $D_1$.  This means that any non-triangle and non-quadrilateral faces must be common between $D_1$ and $D_2$.  \hlfix{These common faces are irrelevant for the statement}{then why in the statement of thm?}, and so by triangulating any common non-triangle faces, we may assume that $D_1$ and $D_2$ have only quadrilateral and triangle faces.
